Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-01-10 paper.Rnw
Fixes to the WA xval presentation

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
%  - ADNI-1 or ADNI1, standardize

@@

\paragraph{Evaluation method}  
Evaluating the accuracy of hippocampal subfields for T1 images at
$0.9mm^3$-isotropic voxels is inherently ill-defined since there are no manual
protocols for segmentation at this resolution to use as a reference. Instead, we
can evaluate the {\it precision} with which \mb{} produces hippocampal subfields
that correspond to the segmentation protocol used by the given atlas library
images.  Specifically, we compute the relative percent error in volume of the
\mb{}-produced hippocampal subfields with respect to the full-resolution
Winterburn atlas segmentations. In addtion, by directly resampling the
Winterburn atlas segmentations to $0.9mm^3$ voxels (using standard
nearest-neighbour image resampling techniques) we obtain a scaled down version
of the labels preserving the original segmentation protocol within the limits of
error from rounding and interpolation. The relative percent error in volume of
these labels provides a baseline to compare \mb{} segmentations against.

@@

In this experiment the \mb{} segmentation of hippocampal subfields is assessed
through a leave-one-out validation in which resampled images ($0.9mm^3$
voxels) of each Winterburn atlas subject are segmented using the remaining
high-resolution Winterburn atlas subjects' images as atlases.

Figure \ref{fig:WAval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subregion across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subregions are: 
  \Sexpr{mean(CA1)}  $\pm$\Sexpr{sd(CA1)} for the CA1 region,  
  \Sexpr{mean(CA23)} $\pm$\Sexpr{sd(CA23)} for the CA2/CA3 regions,  
  \Sexpr{mean(CA4DG)}$\pm$\Sexpr{sd(CA4DG)} for the CA4/DG regions,  
  \Sexpr{mean(Sub)}  $\pm$\Sexpr{sd(Sub)} for the Subiculum region, and
  \Sexpr{mean(SRLM)} $\pm$\Sexpr{sd(SRLM)} for the moleculare regions.

Bland-Altman plots comparing \mb{} to resampled subregion volumes reveal a
proportional bias towards underestimating volumes in all subregions (Figure
\ref{fig:WAxval-Bland-Altman}). Prominent fixed biases in the volumes of the
CA4/DG and SR/SL/SM subregions are congruent with the lower similarity scores
seen in these regions.

@@

caption="Overlap similarity results for the each of the subfields of the hippocampus. 
Simulated overlap similarity results are also given for resampled labels that were 
translated by one voxel (i.e.: $0.9mm$ in all directions), scaled by 0.99, and scaled
by 1.01 (representing a 1\\% shrinkage or scale in all three cardinal directions).
Values are given as mean Dice's Kappa coefficient \\pm stddev."
latex(wa.perturb.table, #file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      label="tab:WAxval-perturb-kappa", rowname=NULL)
@

\begin{figure}
<<fig:WAxval-Bland-Altman>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv')
winterburn = read.csv('data/winterburn.volumes.csv')

levels(wa.n.all.vols$variable) <-list(
  "CA1"="X1", "CA1"="X101", "Subiculum"="X2","Subiculum"="X102", 
  "CA4/DG"="X4","CA4/DG"="X104", "CA2/CA3"="X5","CA2/CA3"="X105", 
  "SR/SL/SM"="X6","SR/SL/SM"="X106")

wa.n.all.vols$diff = wa.n.all.vols$'0.9' - wa.n.all.vols$value
wa.n.all.vols$mean = (wa.n.all.vols$'0.9' + wa.n.all.vols$value)/2

wa.n.xval.bland.limits = ddply(wa.n.all.vols,c("variable"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})
ggplot(wa.n.all.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean of resampled and \\mb subregion volume ' (mm^3))) + 
  ylab(expression('Difference between manual and \\mb subregion volume ' (mm^3))) + 
  scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{\textbf{Bland-Altman plots comparing unilateral hippocampal subregion volume
across all folds of Winterburn atlas cross validation.} The overall mean
difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by dashed
horizontal lines. Linear fit lines are shown. Note, points below the mean
difference indication overestimation of the volume with respect to the resampled
volume, and vice versa. \label{fig:WAxval-Bland-Altman}}
\end{figure}

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}.  Table
\ref{tab:subfield-segmentation-kappa} compares segmentation accuracy from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subregions are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel.  In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores.  Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases which, based on our findings with whole
hippocampal segmentation, would lead to improved overlap similarity.
Nevertheless, the advantage of evaluating \mb{} on standard 3T T1-weighted
resolution MR images with a publically available atlas library is that our
results reflect typical usage scenarios of researchers and clinicians.

@@

In conclusion, we have presented a flexible \multiatlas{}-based framework that has
considerable advantages over other methods since only a small number of atlases
is required to initialize the algorithm. We demonstrate that our method works
robustly over hippocampal definitions, different disease populations, and
different acquisition types. Finally, we also demonstrate that accurate
identification of the hippocampal subfields is possible. 
% TODO: say that we have shown the viability of template generation, and as such
% it may also serve as a useful addition to existing segmentation techniques

@@

ggplot(subset(stats.tests, direction > 0 & stat == "Variance"), 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases))) + 
  #facet_grid( . ~ stat) + 
  geom_line(size=0.5) +
  geom_point(size=2) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  #  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "Variability (p)") +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  theme(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb{} vs. \multiatlas{} segmentation 
  with respect to \snt{} segmentations.}
  Variance of segmentation accuracy between \mb{} and \multiatlas{} segmentation is
  computed for each subject across all ten rounds of validation. Shown on the
  y-axis is the p-value from a t-test comparing the distribution of variances at
  each parameter setting (atlas/template library size). 
  Only points where \mb{} mean variability is lower than \multiatlas{} are shown.
  Dashed lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-SNT-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
%  - remove 'accuracy' unless specifically talking relative to ground truth (and
%    even then, qualify)
%  - overlap/similarity --> agreement? choose one. 
%  - check that chart labels are sensible 
%  - in particular, check that Experiment 5 match Experiment 1
%  - fix figure 2 schematic language for the experiments (not 2a 2b). 
%  - ADNI-1 or ADNI1, standardize
%  - Don't use "Experiment X", instead use some short form description, and then
%  reference (Experiment X).

@@

\paragraph{Evaluation method}  
Evaluating the accuracy of hippocampal subfields for T1 images at
$0.9mm^3$-isotropic voxels is inherently ill-defined since there are no manual
protocols for segmentation at this resolution to use as a reference. Instead, we
can evaluate the reliability, or {\it precision}, with which \mb{} produces
hippocampal subfields that correspond to the segmentation protocol used by the
given atlas library images. By directly resampling the Winterburn atlas
segmentations to $0.9mm^3$ voxels (using standard nearest-neighbour image
resampling techniques) we obtain a scaled down version of the labels which
preserve the original segmentation protocol within the limits of error from
rounding and interpolation. Therefore, using the resampled Winterburn
segmentations as definitive for the $0.9mm^3$ resolution we evaluate reliability
of \mb{} segmentations using using DSC overlap scores and evaluate consistency
across the range of hippocampal sizes using Bland-Altman plots of subfield
volumes.

@@

In this experiment the \mb{} segmentation of hippocampal subfields is assessed
through a leave-one-out validation in which resampled images ($0.9mm^3$ voxels)
of each Winterburn atlas subject are segmented using the remaining
high-resolution Winterburn atlases.

Figure \ref{fig:WAval-kappa} shows the overlap similarity scores between the
\mb{} segmentations and the resampled Winterburn atlases for each hippocampal
subregion across all subjects and folds of the validation. Mean and standard
deviation DSC scores of the subregions are shown in Table
\ref{tab:WAxval-perturb-kappa}, along with DSC scores for the resampled atlas
segmentations when perturbed slightly and compared to the originals.  We find
that the CA4/DG subregion shows the highest mean DSC score of
\Sexpr{mean(CA4DG)} $\pm$\Sexpr{sd(CA4DG)}, followed by the Subiculum and CA1
subregions having scores of \Sexpr{mean(CA1)} $\pm$\Sexpr{sd(CA1)} and
\Sexpr{mean(Sub)} $\pm$\Sexpr{sd(Sub)}, respectively.  Both the CA4/DG and
molecular regions score below 0.5.  These scores may seem low but not when taken
in context and compared to existing (semi-)automated methods (see Discussion).

\begin{figure}
<<fig:WAxval-Bland-Altman>>=
wa.n.vol = read.csv('data/winterburn.xval.volumes.csv')
winterburn = read.csv('data/winterburn.volumes.csv')
subfield_labels <-list(
  "CA1"="X1", "CA1"="X101", 
  "CA2/CA3"="X5","CA2/CA3"="X105", 
  "CA4/DG"="X4","CA4/DG"="X104",
  "Subiculum"="X2","Subiculum"="X102",
  "SR/SL/SM"="X6","SR/SL/SM"="X106")
hemispheres <- list(
  "Right"="X1","Left"="X101", 
  "Right"="X5","Left"="X105", 
  "Right"="X4","Left"="X104",
  "Right"="X2","Left"="X102",
  "Right"="X6","Left"="X106")

levels(wa.n.all.vols$subject) <- list(
  "1"="anusha", 
  "2"="jon",
  "3"="mallar",
  "4"="sofia",
  "5"="nancy")

# mean per subfield/hemisphere across all 4 validation folds
wa.n.mean.vols = ddply(wa.n.all.vols, c("variable", "subject", "res"), function(df) {
  data.frame(mb = mean(df$value),
             high  = mean(df$'0.3'),
             low   = mean(df$'0.9'))
})
wa.n.mean.vols$diff = wa.n.mean.vols$low - wa.n.mean.vols$mb
wa.n.mean.vols$mean = (wa.n.mean.vols$low + wa.n.mean.vols$mb)/2

wa.n.mean.vols$hemisphere <- wa.n.mean.vols$variable
levels(wa.n.all.vols$variable)   <- subfield_labels
levels(wa.n.mean.vols$variable)  <- subfield_labels
levels(wa.n.all.vols$hemisphere) <- hemispheres

wa.n.xval.bland.limits = ddply(wa.n.mean.vols,c("variable"),function(df) {
  data.frame(y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T))
})

wa.n.xval.bland.cor = ddply(wa.n.mean.vols,c("variable"),function(df) {
  cor.t = cor.test(df$low,df$mb)
  data.frame(
    pearson = sprintf("r=%0.2f, t=%0.2f, p < %0.2f", 
                cor.t$estimate, cor.t$statistic, cor.t$p.value)
  )
})

ggplot(wa.n.mean.vols, aes(x=mean, y=diff)) + 
  facet_wrap(~variable, ncol=3, scales='free_x') + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=2, aes(colour=hemisphere, shape=subject)) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=wa.n.xval.bland.limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", 
             data = wa.n.xval.bland.limits, hjust=1.1,vjust=-0.8, size=2.5) + 
  geom_text(aes(y = Inf, x=-Inf, label=pearson), colour="black", 
             data = wa.n.xval.bland.cor, hjust=-0.1,vjust=2.5, size=2.5) + 
  xlab(expression('Mean of resampled and MAGeT-Brain subregion volume ' (mm^3))) + 
  ylab(expression('Difference of manual and MAGeT-Brain subregion volume ' (mm^3))) + 
  #scale_x_continuous(limits=c(0,1000)) + 
  #scale_color_discrete(name="") + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@
  \caption{\textbf{Bland-Altman plots comparing unilateral hippocampal subregion volume
across all folds of Winterburn atlas cross validation.} The overall mean
difference in volume, and limits of agreement ($\pm 1.96SD$) are shown by dashed
horizontal lines. Linear fit lines are shown. Note, points below the mean
difference indication overestimation of the volume with respect to the resampled
volume, and vice versa. \label{fig:WAxval-Bland-Altman}}
\end{figure}

We find that in each subfield, \mb{} shows a proportional bias towards
underestimating volumes in all subregions (Figure
\ref{fig:WAxval-Bland-Altman}).  Prominent fixed biases in the volumes of the
CA4/DG and SR/SL/SM subregions are congruent with the lower similarity scores
seen in these regions.

@@

caption="Overlap similarity results for the each of the subfields of the hippocampus. 
Simulated overlap similarity results are also given for resampled labels that were 
translated by one voxel (i.e.: $0.9mm$ in all directions), scaled by 0.99, and scaled
by 1.01 (representing a 1\\% shrinkage or scale in all three cardinal directions).
Values are given as mean Dice's Kappa coefficient $\\pm$ stddev."
latex(wa.perturb.table, file="", size="scriptsize",landscape=FALSE, ctable=F,
      caption=caption, caption.loc = c("top"),
      col.just=rep('c',5),  
      label="tab:WAxval-perturb-kappa", rowname=NULL)
@

@@

Finally, we have provided evidence that using the Winterburn high-resolution
hippocampal subfield atlases \citep{Winterburn2013} our algorithmic framework is
appropriate for the segmentation of hippocampal subfields in standard
T1-weighted data. Subfield segmentation is a burgeoning topic in the literature
although very few automated methods are available for the segmentation of 3T
data \citep{Yushkevich2009,Yushkevich2010,VanLeemput2009}.  Table
\ref{tab:subfield-segmentation-kappa} compares segmentation accuracy from some
of these methods and \mb{}. The overlap DSC scores for \mb{} subregions are
notably lower but a direct comparison of overlap values must be done cautiously.
In the present work, our overlap scores are computed on $0.9mm$-isotropic voxel
resolution images, whereas \citet{Yushkevich2010} uses focal
$0.4\times0.5\times2.0mm$ voxel resolution images, and \citet{VanLeemput2009}
use supersampled $0.380mm$-isotropic voxel resolution images. The larger voxel
images we use necessarily entail a greater change in DSC for each incorrectly
labelled voxel.  In addition, our automated segmentations are compared to manual
segmentations resampled from $0.3mm$-isotropic voxel labels; the resampling
process inevitably introduces noise which may lower overlap scores.  Lastly, as
our method is aimed specifically at situations when manually produced atlases
are scarce, in our cross validation we are forced to use three rather than all
five of the Winterburn atlases which, based on our findings with whole
hippocampal segmentation, would lead to improved overlap similarity.
Nevertheless, the advantage of evaluating \mb{} on standard 3T T1-weighted
resolution MR images with a publically available atlas library is that our
results reflect typical usage scenarios of researchers and clinicians. 
%TODO: mention that DSC by it's nature penalizes structures with high surface
%area-to-vol ratios, so necessarily penalise all subfield errors.
%TODO: discuss perturbed segmentation kappa values in relation to MAGeT error

@@

In conclusion, we have demonstrated the viability of leveraging a small
number of input atlases to bootstrap a large template library and thereby
improve segmentation reliability when using \multiatlas{} methods.  We
demonstrated that this method works robustly over hippocampal definitions,
different disease populations, and different acquisition types. Finally, we also
demonstrate that accurate identification of the hippocampal subfields in
standard 3T T1-weighted images is possible.

@@

  # test statistic is positive if MA mean is larger than MB
  # i.e. that MA has a greater mean of variances/SDs
  # i.e. that MB has a smaller mean, and so is doing "better"
  data.frame(
    stat      = c("Variance", "Standard Deviation"),
    statistic = c(t.var$statistic, t.sd$statistic),
    p.value   = c(t.var$p.value, t.sd$p.value),
    direction = c(diff_mean_var, diff_mean_sd)
  )
})
  y-axis is the p-value from a t-test comparing the distribution of variances at
  each parameter setting (atlas/template library size). 
  Only points where \mb{} mean variability is lower than \multiatlas{} are shown.
  Dashed lines indicate a p-value of 0.05 and 0.01.}
  \label{ADNI1-SNT-xval-variability}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
